---
title: "Running the EvoDemo-Hyper Matrix Population Model in R"
author: "Timothée Bonnet"
date: "2024-10-28"
output: 
  html_document: 
    toc: yes
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    theme: cerulean
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code is associated with the manuscript "Toward a unified approach to modeling adaptation among demographers and evolutionary ecologists" submitted to Methods in Ecology and Evolution (MEE) by Joanie Van de Walle, Jimmy Garnier, Timothée Bonnet and Stephanie Jenouvrier.

# Set-up


R-session and computer state when compiling this file:

```{r, eval=TRUE}
sessionInfo()
```


Load all MPM simulation functions:

```{r, eval=TRUE}
source("MPM_Functions.R")
```


# Define species

We define 5 species as a collection of vital rates and life-history properties.

```{r, eval=FALSE}
species_info <- data.frame(species=1:5,
                           fecundity = c(10.9868, 4.9958, 0.9987 ,0.3004 , 0.2286),
                           fertility = c(10.9868/0.0965, 4.9958/0.25, 0.9987/0.385 ,0.3004/0.505 , 0.2286/0.8),
                           adult_survival = c(0.0365, 0.2, 0.8, 0.93, 0.95), 
                           juvenile_survival = c(0.0965, 0.25, 0.385, 0.505, 0.8),
                           maturation_rate = c(0.9, 0.572, 0.4, 0.3, 0.07), 
                           generation_time = c(2.0476, 2.3697, 6.2983, 15.8066, 23.7847))

```



