---
title: "Running the EvoDemo-Hyper Matrix Population Model in R"
author: "Timothée Bonnet"
date: "2024-10-28"
output: 
  html_document: 
    toc: yes
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    theme: cerulean
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code is associated with the manuscript "Toward a unified approach to modeling adaptation among demographers and evolutionary ecologists" submitted to Methods in Ecology and Evolution (MEE) by Joanie Van de Walle, Jimmy Garnier, Timothée Bonnet and Stephanie Jenouvrier.

# Set-up

Packages that are needed to run the MPM function:

```{r}
library(pracma) # for data wrangling and plotting
library(gsignal) # for population growth rate
library(purrr) # for reformatting output
```


Packages to visualise results:

```{r}
library(tidyverse)
```




R-session and computer state when compiling this file:

```{r, eval=TRUE}
sessionInfo()
```


Load all MPM simulation functions:

```{r, eval=TRUE}
source("MPM_Functions.R")
```


# Define species

We define 5 species as a collection of vital rates and life-history properties.

```{r}
species_info <- data.frame(species=1:5,
                           fecundity = c(10.9868, 4.9958, 0.9987 ,0.3004 , 0.2286),
                           fertility = c(10.9868/0.0965, 4.9958/0.25, 0.9987/0.385 ,0.3004/0.505 , 0.2286/0.8),
                           adult_survival = c(0.0365, 0.2, 0.8, 0.93, 0.95), 
                           juvenile_survival = c(0.0965, 0.25, 0.385, 0.505, 0.8),
                           maturation_rate = c(0.9, 0.572, 0.4, 0.3, 0.07), 
                           generation_time = c(2.0476, 2.3697, 6.2983, 15.8066, 23.7847))

```

# Demonstration

## An example

A single function is called to run the EvoDemo-Hyper Matrix Population Model (MPM). 

We start by demonstrating the use of the function for species 1, using default parameters as used in the manuscript (this should take a few seconds):

```{r}
output1 <- MPM(h2 = 0.2, Vp=1, nind = 100, Beta = 0.15,
               Fe = species_info$fecundity[1], 
               SA = species_info$adult_survival[1],
               SJ = species_info$juvenile_survival[1],
               S0 = species_info$juvenile_survival[1],
               Y = species_info$maturation_rate[1])

```

Note that we use the same survival probability, `species_info$juvenile_survival[1]`, for `SJ` (survival probability for juvenile individuals, who have not matured yet) and `S0` (survival probability on the birth year). It is possible to use different probabilities for those two rates, as was done in the appendix of the manuscript.

The output contains lots of information. 

```{r}
str(output1)
```


To start with, we can visualise population size:

```{r}
plot(output1$properties_per_year[,c("year", "N")])
```

We can extract the distribution of phenotypes through time in `$pheno_distribution`:

```{r}
pheno_distribution <- output1$pheno_distribution
```

We can visualise the number of individuals expected in each phenotypic class through 100 years of simulation:

```{r}
pheno_distribution %>% ggplot(aes(x=year, y=density, colour=as.factor(phenotypic_midvalue))) + geom_line()
```

The distribution of phenotypes every 20 years:

```{r}
pheno_distribution %>% filter(year %in% seq( from=1, to =101, by=20)) %>%
  ggplot(aes(x=phenotypic_midvalue, y=density, fill=as.factor(phenotypic_midvalue))) +
  geom_col() +
  facet_wrap(~year, scales = "free_y", labeller = ) + #######################################################
  theme_bw() +
  theme(legend.position = "none")
```


The mean breeding values among adults and juveniles through time:

```{r}
BV <- output1$properties_per_year[,c("year", "M_EBVSA", "M_EBVSJ")]
colnames(BV) <- c("year", "adult", "juvenile")
BV <- BV %>% pivot_longer(cols=2:3, values_to = "mean_breeding_value", names_to = "age_class")

ggplot(BV, aes(x=year, y=mean_breeding_value, color=age_class)) + 
  ylim(c(-4, +4)) + 
  geom_line() + theme_bw() 
```

## Options

### Phenotypic classes

We can change the number of phenotypic and breeding value classes considered (`g=`). The default here was 20, which takes about 5 seconds on this machine. With only 10 classes the algorithm takes about 1 second:

```{r}
output2 <- MPM(g = 10,
               Fe = species_info$fecundity[1], 
               SA = species_info$adult_survival[1],
               SJ = species_info$juvenile_survival[1],
               S0 = species_info$juvenile_survival[1],
               Y = species_info$maturation_rate[1])

```


```{r}
output2$pheno_distribution %>% 
  filter(year %in% seq( from=1, to =101, by=20)) %>%
  ggplot(aes(x=phenotypic_midvalue, y=density, fill=as.factor(phenotypic_midvalue))) +
  geom_col() +
  facet_wrap(~year, scales = "free_y", labeller = ) + 
  theme_bw() +
  theme(legend.position = "none")
```

Using 40 classes takes much more RAM and CPU-time (On this machine, around 20 minutes allocating 14 CPUs), much more than the MATLAB version of the algorithm.

The initial phenotype has a mean of zero. We set the range of phenotypes considered around that value with the parameter `xp=`. The default is 4. We can increase it to 8 so give more space for the trait to evolve:

```{r}
output3 <- MPM(g = 20, 
               xp = 8,
               Fe = species_info$fecundity[1], 
               SA = species_info$adult_survival[1],
               SJ = species_info$juvenile_survival[1],
               S0 = species_info$juvenile_survival[1],
               Y = species_info$maturation_rate[1])
```


```{r}
output3$pheno_distribution %>% 
  filter(year %in% seq( from=1, to =101, by=20)) %>%
  ggplot(aes(x=phenotypic_midvalue, y=density, fill=as.factor(phenotypic_midvalue))) +
  geom_col() +
  facet_wrap(~year, scales = "free_y", labeller = ) + 
  theme_bw() +
  theme(legend.position = "none")
```

Given the broader range there was not enough time to see much change. We can increase the number of timesteps from 100 to 500:

```{r}
output4 <- MPM(g = 20, 
               xp = 8,
               timeA = 500,
               Fe = species_info$fecundity[1], 
               SA = species_info$adult_survival[1],
               SJ = species_info$juvenile_survival[1],
               S0 = species_info$juvenile_survival[1],
               Y = species_info$maturation_rate[1])
```

```{r}
output4$pheno_distribution %>% 
  filter(year %in% seq( from=1, to =501, by=50)) %>%
  ggplot(aes(x=phenotypic_midvalue, y=density, fill=as.factor(phenotypic_midvalue))) +
  geom_col() +
  facet_wrap(~year, scales = "free_y", labeller = ) + 
  theme_bw() +
  theme(legend.position = "none")
```

Eventually the phenotypic distribution hits the boundary of the highest possible phenotype, which makes the simulation unreliable from that point in time (around year 350 here).
Also, note the absurd exponential population growth, evidenced by the y-axis. This is expected as the model consider only directional selection and does not contain density dependence.

### Heritability and selection

We can check that without selection there is no phenotypic or genetic change:

```{r}
output_nobeta <- MPM(g = 20, 
               xp = 4,
               timeA = 100,
               h2 = 0.2,
               Beta = 0,
               nind = 100,
               Fe = species_info$fecundity[4], 
               SA = species_info$adult_survival[4],
               SJ = species_info$juvenile_survival[4],
               S0 = species_info$juvenile_survival[4],
               Y = species_info$maturation_rate[4])

```

```{r}
output_nobeta$pheno_distribution %>% 
  filter(year %in% seq( from=1, to =101, by=10)) %>%
  ggplot(aes(x=phenotypic_midvalue, y=density, fill=as.factor(phenotypic_midvalue))) +
  geom_col() +
  facet_wrap(~year, scales = "free_y", labeller = ) + 
  theme_bw() +
  theme(legend.position = "none")

```

```{r}
BV <- output_nobeta$properties_per_year[,c("year", "M_EBVSA", "M_EBVSJ")]
colnames(BV) <- c("year", "adult", "juvenile")
BV <- BV %>% pivot_longer(cols=2:3, values_to = "mean_breeding_value", names_to = "age_class")

ggplot(BV, aes(x=year, y=mean_breeding_value, color=age_class)) + 
  ylim(c(-4, +4)) + 
  geom_line() + theme_bw() 
```

Note however that the population size tends to increase after an initial drop:

```{r}
plot(output_nobeta$properties_per_year[,c("year", "N")])
```

The asymptotic population growth rate is constant as there is no adaptation, but it is very slightly greater than 1:

```{r}
plot(output_nobeta$properties_per_year[,c("year", "lambda")])
```


When there is selection but no heritability (we set the heritability to 0.001 as it cannot be exactly 0 or 1):
 
```{r}
output_smallh2 <- MPM(g = 20, 
               xp = 4,
               timeA = 100,
               h2 = 0.01,
               Beta = 0.15,
               nind = 100,
               Fe = species_info$fecundity[4], 
               SA = species_info$adult_survival[4],
               SJ = species_info$juvenile_survival[4],
               S0 = species_info$juvenile_survival[4],
               Y = species_info$maturation_rate[4])

```

We get a warning message. The dynamic is not necessarily wrong, but maybe not what you expected:

population size increases more:

```{r}
plot(output_smallh2$properties_per_year[,c("year", "N")])
```

Due to an initial increase than stabilization of lambda:

```{r}
plot(output_smallh2$properties_per_year[,c("year", "lambda")])
```

What we see is a little bit of adaptation, eroding away the additive genetic variance:

```{r}
plot(output_smallh2$properties_per_year[,c("year", "Va")])
```

This erosion may be seen as realistic, but also as an artefact of the algorithm. We do not model genes explicitly, so we do not have a clear prediction of what should happen to Va.

```{r}
output_smallh2$pheno_distribution %>% 
  filter(year %in% seq( from=1, to =101, by=10)) %>%
  ggplot(aes(x=phenotypic_midvalue, y=density, fill=as.factor(phenotypic_midvalue))) +
  geom_col() +
  facet_wrap(~year, scales = "free_y") + 
  theme_bw() +
  theme(legend.position = "none")


```
