---
title: "Running the Individual Based Model"
author: "Timothée Bonnet"
date: "2024-10-28"
output: 
  html_document: 
    toc: yes
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    theme: cerulean
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

This code is associated with the manuscript "Toward a unified approach to modeling adaptation among demographers and evolutionary ecologists" submitted to Methods in Ecology and Evolution (MEE) by Joanie Van de Walle, Jimmy Garnier, Timothée Bonnet and Stephanie Jenouvrier.

# Set-up

Load the packages needed to run the code:

```{r}
library(doParallel) # to run simulations multicore
library(ggpubr) # graphes
library(popbio) # population matrix calculations
library(tidyverse) # data wrangling
library(viridis) # colours for graphes
```

R-session and computer state when compiling this file:

```{r, eval=TRUE}
sessionInfo()
```


Load all IBM simulation and post-treatment functions:

```{r, eval=TRUE}
source("IBM_Function.R")
```


# Define species

We define 5 species as a collection of vital rates and life-history properties.

```{r, eval=FALSE}
species_info <- data.frame(species=1:5,
                           fecundity = c(10.9868, 4.9958, 0.9987 ,0.3004 , 0.2286),
                           fertility = c(10.9868/0.0965, 4.9958/0.25, 0.9987/0.385 ,0.3004/0.505 , 0.2286/0.8),
                           adult_survival = c(0.0365, 0.2, 0.8, 0.93, 0.95), 
                           juvenile_survival = c(0.0965, 0.25, 0.385, 0.505, 0.8),
                           maturation_rate = c(0.9, 0.572, 0.4, 0.3, 0.07), 
                           generation_time = c(2.0476, 2.3697, 6.2983, 15.8066, 23.7847))

```

Check expected asymptotic growth rate and stage structure

```{r, eval=FALSE}
species_info$lambda <- NA
species_info$juv_prop <- NA

for (i in 1:nrow(species_info))
{
  leslie_pop <- matrix(c(species_info$juvenile_survival[i] * ( 1 - species_info$maturation_rate[i]),
         species_info$fecundity[i],
         species_info$juvenile_survival[i] * species_info$maturation_rate[i],
         species_info$adult_survival[i]),
       nrow = 2, byrow = TRUE)


  species_info$lambda[i] <- lambda(leslie_pop)
  species_info$juv_prop[i] <- stable.stage(leslie_pop)[1]
}

write_csv(species_info, file = "species_info.csv")
```

```{r}
species_info <- read.csv("species_info.csv")

species_info

```




# Run the IBM across species and vital rates

We run the IBM one vital rate at the time, looping over the species. 

In order to speed up computation we use parallel computing for the inner for-loop. Assuming you have 4 independent cores available to run the IBM, you can start the virtual cluster with:

```{r, eval=FALSE}
cl <- makeCluster(4)
registerDoParallel(cl)
```

If you have more cores available, for instance 20 you would use:

```{r, eval=FALSE}
cl <- makeCluster(20)
registerDoParallel(cl)
```

After running one of the for-loops below you should close the virtual cluster with:

```{r, eval=FALSE}
stopCluster(cl)
```

This will shut down the background R-sessions.


If you do not want to use parallel computing, you should replace `foreach (j = 1:100) %dopar% ` with `for (j in 1:100) `.


## Juvenile survival selection 


```{r, eval=FALSE}
source("FunctionSimulsIPMQG.R")

species_info <- read.csv("species_info.csv")

VP <- 1
heritability <- 0.2
beta_s_j <- 0.15
init_pop_size <- 2*100 # multiply by 2 to account for males
nb_years <- 100

K <- 2000 # safeguard carrying capacity to prevent very long simulations

for (i in 1:nrow(species_info))
{
  survival_j <- species_info$juvenile_survival[i]
  survival_a <- species_info$adult_survival[i]
  maturation <- species_info$maturation_rate[i]
  fecundity <- species_info$fertility[i]*2 # multiply by 2 to account for males # in the case of Juv survival we need to use Fertility instead of Fecundity
  init_prop_juv <- species_info$juv_prop[i]

  foreach (j = 1:100) %dopar%
  {
    if(! paste0("pop_sel_species_s_j_",i, "_rep_",j) %in% list.files("simulations") )
      {
      pop_sel <- main_simul(init_pop_size=init_pop_size, 
                         start = 1, end = nb_years, 
                         fecundity = fecundity, VP = VP, VA = heritability*VP,
                         maturation_rate = maturation,
                         surv_j = survival_j, surv_a = survival_a, beta_s_j = beta_s_j, init_prop_juv=init_prop_juv, 
                         K=K)
      save(pop_sel, file = paste0("simulations/pop_sel_species_s_j_",i, "_rep_",j))
    }
  print(j)
  }
}

```

### Post-hoc treatment of simulations

Produce demographic data. This is quite slow (and could probably be optimized a lot).

```{r, eval=FALSE}

for (i in 1:nrow(species_info))
{
  foreach (j = 1:100) %dopar%
  {
    if(paste0("pop_sel_species_s_j_",i, "_rep_",j) %in% list.files("simulations") )
      if(!paste0("pop_data_species_s_j_",i, "_rep_",j, ".csv") %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_s_j_",i, "_rep_",j))
        pop_data <- produce_pop_data(pop_sel)
        write_csv(pop_data, file = paste0("simulations/pop_data_species_s_j_",i, "_rep_",j, ".csv"))
      }
    print(j)
  }
}

pop_data_all_s_j <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    pop_data <- read.csv(paste0("simulations/pop_data_species_s_j_",i, "_rep_",j, ".csv"))
    pop_data$species <- i
    pop_data$replicate <- j
    pop_data_all_s_j <- bind_rows(pop_data_all_s_j, pop_data)
  }
}
write.csv(pop_data_all_s_j, "simulations/pop_data_all_s_j.csv")

```

Produce data on average breeding values by cohort or by year:

```{r, eval=FALSE}

all_pop_s_j <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_s_j_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_s_j_",i, "_rep_",j))
        pop_evol <- produce_evolution_data(pop_sel, minyear = 1)
        pop_evol$species <- i
        pop_evol$replicate <- j
        all_pop_s_j <- bind_rows(all_pop_s_j, pop_evol)
      }
  }
}
write.csv(all_pop_s_j, file = "simulations/pop_evol_s_j.csv")

# by year evolution
all_pop_by_s_j <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_s_j_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_s_j_",i, "_rep_",j))
        pop_evol <- produce_evolution_data_byyear(pop_sel, minyear = 1)
        pop_evol$species <- i
        pop_evol$replicate <- j
        all_pop_by_s_j <- bind_rows(all_pop_by_s_j, pop_evol)
    }
  }
}

write.csv(all_pop_by_s_j, file = "simulations/pop_evol_by_s_j.csv")

```



## Adult survival selection

```{r, eval=FALSE}
source("FunctionSimulsIPMQG.R")

library(tidyverse)
species_info <- read.csv("species_info.csv")

VP <- 1
heritability <- 0.2
beta_s_a <- 0.15
init_pop_size <- 2*100 # multiply by 2 to account for males
nb_years <- 100

K <- 5000 #safeguard to prevent very long simulations

for (i in 1:nrow(species_info))
{
  survival_j <- species_info$juvenile_survival[i]
  survival_a <- species_info$adult_survival[i]
  maturation <- species_info$maturation_rate[i]
  fecundity <- species_info$fecundity[i]*2 # multiply by 2 to account for males
  init_prop_juv <- species_info$juv_prop[i]

  for (j in 1:100)
  {
    if(! paste0("pop_sel_species_sa_",i, "_rep_",j) %in% list.files("simulations") )
      {
      pop_sel <- main_simul(init_pop_size=init_pop_size, 
                         start = 1, end = nb_years, 
                         fecundity = fecundity, VP = VP, VA = heritability*VP,
                         maturation_rate = maturation,
                         surv_j = survival_j, surv_a = survival_a, beta_s_a=beta_s_a,
                         init_prop_juv=init_prop_juv, 
                         K=K)
      save(pop_sel, file = paste0("simulations/pop_sel_species_sa_",i, "_rep_",j))
    }
  print(j)
  }
}

```

```{r, eval=FALSE}

for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_sa_",i, "_rep_",j) %in% list.files("simulations") )
      if(!paste0("pop_data_species_sa_",i, "_rep_",j, ".csv") %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_sa_",i, "_rep_",j))
        pop_data <- produce_pop_data(pop_sel)
        write_csv(pop_data, file = paste0("simulations/pop_data_species_sa_",i, "_rep_",j, ".csv"))
      }
    print(j)
  }
}

pop_data_all_s_a <- tibble()

for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    pop_data <- read.csv(paste0("simulations/pop_data_species_sa_",i, "_rep_",j, ".csv"))
    pop_data$species <- i
    pop_data$replicate <- j
    pop_data_all_s_a <- bind_rows(pop_data_all_s_a, pop_data)
  }
}

write.csv(pop_data_all_s_a, "simulations/pop_data_all_s_a.csv")
```


```{r, eval=FALSE}
all_pop_s_a <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_sa_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_sa_",i, "_rep_",j))
        pop_evol <- produce_evolution_data(pop_sel, minyear = 1)
        pop_evol$species <- i
        pop_evol$replicate <- j
        all_pop_s_a <- bind_rows(all_pop_s_a, pop_evol)
    }
  }
}

write.csv(all_pop_s_a, file = "simulations/pop_evol_s_a.csv")

#by year evolution
all_pop_by_s_a <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_sa_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_sa_",i, "_rep_",j))
        pop_evol <- produce_evolution_data_byyear(pop_sel, minyear = 1)
        pop_evol$species <- i
        pop_evol$replicate <- j
        all_pop_by_s_a <- bind_rows(all_pop_by_s_a, pop_evol)
    }
  }
}

write.csv(all_pop_by_s_a, file = "simulations/pop_evol_by_s_a.csv")

```


## Fertility selection

```{r, eval=FALSE}
source("FunctionSimulsIPMQG.R")

library(tidyverse)
species_info <- read.csv("species_info.csv")

VP <- 1
heritability <- 0.2
beta_f <- 0.15*2
init_pop_size <- 2*100 # multiply by 2 to account for males
nb_years <- 100

K <- 5000 #safeguard to prevent very long simulations

for (i in 1:nrow(species_info))
{
  survival_j <- species_info$juvenile_survival[i]
  survival_a <- species_info$adult_survival[i]
  maturation <- species_info$maturation_rate[i]
  fecundity <- species_info$fecundity[i]*2 # multiply by 2 to account for males
  init_prop_juv <- species_info$juv_prop[i]

  for (j in 1:100)
  {
    if(! paste0("pop_sel_species_f_",i, "_rep_",j) %in% list.files("simulations") )
      {
      pop_sel <- main_simul(init_pop_size=init_pop_size, 
                         start = 1, end = nb_years, 
                         fecundity = fecundity, VP = VP, VA = heritability*VP,
                         maturation_rate = maturation,
                         surv_j = survival_j, surv_a = survival_a, beta_f=beta_f,
                         init_prop_juv=init_prop_juv, 
                         K=K, repro_model="poisson")
      save(pop_sel, file = paste0("simulations/pop_sel_species_f_",i, "_rep_",j))
    }
  print(j)
  }
}

```

```{r, eval=FALSE}
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_f_",i, "_rep_",j) %in% list.files("simulations") )
      if(!paste0("pop_data_species_f_",i, "_rep_",j, ".csv") %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_f_",i, "_rep_",j))
        pop_data <- produce_pop_data(pop_sel)
        write_csv(pop_data, file = paste0("simulations/pop_data_species_f_",i, "_rep_",j, ".csv"))
      }
    print(j)
  }
}

pop_data_all_f <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    pop_data <- read.csv(paste0("simulations/pop_data_species_f_",i, "_rep_",j, ".csv"))
    pop_data$species <- i
    pop_data$replicate <- j
    pop_data_all_f <- bind_rows(pop_data_all_f, pop_data)
  }
}
write.csv(pop_data_all_f, "simulations/pop_data_all_f.csv")

all_pop_f <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_f_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_f_",i, "_rep_",j))
        pop_evol <- produce_evolution_data(pop_sel, minyear = 1)
        pop_evol$species <- i
        pop_evol$replicate <- j
        all_pop_f <- bind_rows(all_pop_f, pop_evol)
      }
  }
}
write.csv(all_pop_f, file = "simulations/pop_evol_f.csv")

#by year calculation instead of by cohort
all_pop_f_by <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_f_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_f_",i, "_rep_",j))
        pop_evol_by <- produce_evolution_data_byyear(pop_sel, minyear = 1)
        pop_evol_by$species <- i
        pop_evol_by$replicate <- j
        all_pop_f_by <- bind_rows(all_pop_f_by, pop_evol_by)
    }
  }
}

write.csv(all_pop_by_f, file = "simulations/pop_evol_by_f.csv")

```


## Maturity selection

```{r, eval=FALSE}
source("FunctionSimulsIPMQG.R")

library(tidyverse)
species_info <- read.csv("species_info.csv")

VP <- 1
heritability <- 0.2
beta_m <- 0.15
init_pop_size <- 2*100 # multiply by 2 to account for males
nb_years <- 100

K <- 5000 #safeguard to prevent very long simulations

for (i in 1:nrow(species_info))
{
  survival_j <- species_info$juvenile_survival[i]
  survival_a <- species_info$adult_survival[i]
  maturation <- species_info$maturation_rate[i]
  fecundity <- species_info$fecundity[i]*2 # multiply by 2 to account for males
  init_prop_juv <- species_info$juv_prop[i]

  for (j in 1:100)
  {
    if(! paste0("pop_sel_species_m_",i, "_rep_",j) %in% list.files("simulations") )
      {
      pop_sel <- main_simul(init_pop_size=init_pop_size, 
                         start = 1, end = nb_years, 
                         fecundity = fecundity, VP = VP, VA = heritability*VP,
                         maturation_rate = maturation,
                         surv_j = survival_j, surv_a = survival_a, beta_m=beta_m,
                         init_prop_juv=init_prop_juv, 
                         K=K)
      save(pop_sel, file = paste0("simulations/pop_sel_species_m_",i, "_rep_",j))
    }
  print(j)
  }
}

```



```{r, eval=FALSE}
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_m_",i, "_rep_",j) %in% list.files("simulations") )
      if(!paste0("pop_data_species_m_",i, "_rep_",j, ".csv") %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_m_",i, "_rep_",j))
        pop_data <- produce_pop_data(pop_sel)
        write_csv(pop_data, file = paste0("simulations/pop_data_species_m_",i, "_rep_",j, ".csv"))
      }
    print(j)
  }
}

pop_data_all_m <- tibble()

for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    pop_data <- read.csv(paste0("simulations/pop_data_species_m_",i, "_rep_",j, ".csv"))
    pop_data$species <- i
    pop_data$replicate <- j
    pop_data_all_m <- bind_rows(pop_data_all_m, pop_data)
  }
}

write.csv(pop_data_all_m, "simulations/pop_data_all_m.csv")
```

```{r, eval=FALSE}
all_pop_m <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_m_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_m_",i, "_rep_",j))
        pop_evol <- produce_evolution_data(pop_sel, minyear = 1)
        pop_evol$species <- i
        pop_evol$replicate <- j
        all_pop_m <- bind_rows(all_pop_m, pop_evol)
    }
  }
}

write.csv(all_pop_m, file = "simulations/pop_evol_m.csv")

#by year calculation instead of by cohort
all_pop_by_m <- tibble()
for (i in 1:nrow(species_info))
{
  for (j in 1:100)
  {
    if(paste0("pop_sel_species_m_",i, "_rep_",j) %in% list.files("simulations") )
      {
        load(paste0("simulations/pop_sel_species_m_",i, "_rep_",j))
        pop_evol <- produce_evolution_data_byyear(pop_sel, minyear = 1)
        pop_evol$species <- i
        pop_evol$replicate <- j
        all_pop_by_m <- bind_rows(all_pop_by_m, pop_evol)
    }
  }
}

write.csv(all_pop_by_m, file = "simulations/pop_evol_by_m.csv")
```
